---
title: "MRQAP for networks, cognitive social structures and multi-mode data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MRQAP for networks, cognitive social structures and multi-mode data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(MrQAP)
set.seed(20260825)
```

## Why permutation inference

A regression on dyadic data violates the independence assumption in an
obvious way. Every tie sent by actor *i* shares a sender; every tie received
by *j* shares a receiver. Observations are therefore correlated in a
structured way that ordinary standard errors do not know about, and the
resulting p-values are badly anti-conservative.

The quadratic assignment procedure sidesteps the problem instead of
modelling it. Permute the network — relabel the nodes, keeping the structure
intact — refit, and repeat. That gives an empirical null distribution for
each coefficient, built from data with the same dependence structure as the
original but with the hypothesised relationship destroyed. The p-value is
the proportion of permutations that produced a statistic at least as extreme
as the observed one.

`MrQAP` exposes one function, `QAP()`, for all of this.

## A one-mode network

We simulate a 20-node network whose tie strength depends on two dyadic
covariates.

```{r simulate}
n  <- 20
x1 <- matrix(rnorm(n^2), n, n)   # e.g. shared activity
x2 <- matrix(rnorm(n^2), n, n)   # e.g. geographic distance

y <- 1 + 0.8 * x1 - 0.4 * x2 + matrix(rnorm(n^2), n, n)

d <- list(y = y, x1 = x1, x2 = x2)
```

The left-hand side of the formula names the outcome in `data`; the
right-hand side names the predictors.

```{r fit-basic}
fit <- QAP(y ~ x1 + x2, data = d, reps = 200)
fit
```

The table gives, for each coefficient, the proportion of permutations whose
t-value was below the observed one, above it, and above it in absolute
value. The last is the two-sided p-value and is usually what you want.

The intercept is marked `*` because semi-partialling plus provides no test
for it — the procedure works by residualising each predictor on the others,
which has no analogue for a constant.

Coefficients come out where you would expect:

```{r coefs}
round(fit$coefficients, 3)
```

## Getting the results out

Printing is for reading; the extractor methods are for using.

```{r extractors}
coef(fit)

summary(fit)$coefficients          # a data frame, one row per term

as.data.frame(fit)                 # the same, for stacking across models
```

`summary()` prints the same report as the fit itself, with significance
codes added:

```{r summary}
summary(fit)
```

Note how a p-value of zero is shown. A permutation p-value is a proportion
out of `reps` draws, so its resolution is `1/reps` — with 200 permutations
the smallest value distinguishable from zero is 0.005. It is reported as
`<0.005`, not as the `<2e-16` that a conventional coefficient table would
print, because the permutation test cannot support that claim.

### Confidence intervals

`confint()` builds intervals from the *spread* of the permutation
distribution, so they reflect the same dyadic dependence the p-values do:

```{r confint}
confint(fit)
```

It is worth being explicit about what this is not. The permutation
distribution is generated under the null of no effect, so it is centred on
zero — its percentiles form an interval around zero and are not a
confidence interval at all. Only its standard deviation *s* is used, as
`estimate ± 1.96 * s`. That is a normal approximation: for a strongly
skewed coefficient, treat it as indicative.

Compare it with the model-based interval, which assumes the independence
MRQAP exists to avoid — usually narrower, and not consistent with the
reported p-value:

```{r confint-compare}
rbind(permutation = confint(fit)["x1", ],
      model_based = confint(fit$simple_fit)["x1", ])
```

`vcov()`, `fitted()`, `residuals()`, `logLik()`, `AIC()`, `BIC()` and
`nobs()` all describe the *baseline* fit, and carry the same caveat: they
know nothing about the permutation inference.

### Seeing the null distribution

A p-value is a summary of a distribution, and it is worth looking at the
distribution itself. `plot()` draws the permutation draws for each tested
coefficient with the observed statistic marked in red:

```{r plot, fig.width = 7, fig.height = 3.2}
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
plot(fit)
par(op)
```

An effect whose observed statistic sits well outside the histogram is the
picture a small p-value is describing; one sitting inside it is not.

### Predicting

`predict()` gives fitted values on the response scale, and
`type = "matrix"` puts them back into the shape of the network, with `NA`
wherever a cell did not enter the model:

```{r predict}
p_hat <- predict(fit, type = "matrix")
dim(p_hat)
round(p_hat[1:4, 1:4], 2)          # the diagonal is NA, as it was dropped
```

Pass `newdata` — a named list of matrices shaped like the fitted data — to
predict for a different network.

## Choosing the null hypothesis

Two procedures are available.

`nullhyp = "qapspp"` — the default — is Dekker, Krackhardt and Snijders'
semi-partialling plus. For each predictor in turn it regresses that
predictor on all the others, permutes the *residuals*, and refits. This is
the more trustworthy test when the predictors are correlated with each
other, which is the normal case. It costs `reps` permutations per predictor.

`nullhyp = "qapy"` permutes the outcome instead. It is cheaper — `reps`
permutations in total — but is the weaker test under collinearity.

```{r nullhyp}
qy <- QAP(y ~ x1 + x2, data = d, nullhyp = "qapy", reps = 200)

round(rbind(qapspp = fit$abs["t", ], qapy = qy$abs["t", ]), 3)
```

With a single predictor the two coincide, and `QAP()` silently uses `qapy`.

## Families beyond the linear model

Binary ties are the common case. Everything else about the call is the same.

```{r binomial}
eta <- -0.5 + 1.2 * x1
yb  <- matrix(rbinom(n^2, 1, 1 / (1 + exp(-eta))), n, n)

bfit <- QAP(y ~ x1 + x2, data = list(y = yb, x1 = x1, x2 = x2),
            family = "binomial", reps = 200)
round(bfit$coefficients, 3)
```

For binomial models `QAP()` also returns a confusion matrix built by drawing
from the predicted probabilities rather than by rounding them at 0.5, which
gives an honest picture of classification uncertainty:

```{r confusion}
bfit$confusion_matrix$accuracy
```

`family` also accepts `"poisson"`, `"negbin"`, `"zip"` and `"multinom"`.

## Robust standard errors

`use_robust_errors = TRUE` gives heteroskedasticity-consistent standard
errors: the closed-form HC3 for linear models, and the sandwich estimator
for GLMs.

```{r robust}
rfit <- QAP(y ~ x1 + x2, data = d, use_robust_errors = TRUE, reps = 200)
round(rbind(model_based = fit$t, robust = rfit$t), 2)
```

They are not available for mixed models, where `QAP()` warns and falls back
to model-based errors rather than reporting a number that is not what it
claims to be.

## Weights and offsets

`weights` scales each dyad's contribution to the fit; `offset` enters the
linear predictor with its coefficient fixed at one, which is how exposure
enters a count model. Each is a matrix shaped exactly like the dependent
variable:

```{r weights, eval = FALSE}
QAP(ties ~ x1, data = d, family = "poisson",
    offset = log(opportunities))     # rate per opportunity
```

Both work for every family and for both estimators, including GMM, where
they enter the moment conditions directly.

It matters what these do under permutation. A weight and an offset belong
to the *cell* — this sender, this receiver — not to the actors. They
therefore stay in place while the outcome (or, under `qapspp`, the tested
predictor) is relabelled: they are part of the fixed design against which
the null distribution is generated, not quantities that travel with the
permutation. If your weights are a property of the actors rather than of
the dyad, that is a different model and this is not it.

## Restricting permutations to groups

If the network has a known block structure — classrooms within a school,
departments within a firm — permuting freely across it destroys more than
the hypothesised effect. `groups` restricts each permutation to occur within
groups.

```{r groups}
grp <- rep(1:4, each = 5)
QAP(y ~ x1 + x2, data = d, groups = grp, reps = 200)$abs["t", ]
```

## Cognitive social structures

A CSS records each actor's *perception* of the whole network: a 3D array
indexed `[sender, receiver, perceiver]`. `QAP()` detects the extra dimension
and permutes accordingly — one node order applied across all three
dimensions, since sender, receiver and perceiver are the same people.

```{r css}
m  <- 8
a1 <- array(rnorm(m^3), dim = c(m, m, m))
ya <- 0.7 * a1 + array(rnorm(m^3), dim = c(m, m, m))

cfit <- QAP(y ~ x1, data = list(y = ya, x1 = a1), reps = 100)
round(cfit$coefficients, 3)
```

Perceiver-level heterogeneity can be absorbed with a random intercept:

```{r css-re, eval = requireNamespace("lme4", quietly = TRUE)}
QAP(y ~ x1, data = list(y = ya, x1 = a1),
    random_intercept_perceiver = TRUE, reps = 50)$coefficients
```

## Multi-mode data

In two-mode (affiliation) data the rows and columns are *different* kinds of
thing — people and organisations, say. There is no diagonal to exclude,
because an actor cannot have a tie to itself across node sets, and no
reciprocity, because there is no reverse dyad. The permutation must
therefore draw an independent node order for each dimension.

`QAP()` infers this when the dimensions differ in size:

```{r twomode}
nr <- 24; nc <- 8                      # 24 people, 8 organisations
p1 <- matrix(rnorm(nr * nc), nr, nc)
yt <- 0.6 * p1 + matrix(rnorm(nr * nc), nr, nc)

tfit <- QAP(y ~ x1, data = list(y = yt, x1 = p1), reps = 100)
tfit$multi_mode
nrow(model.frame(tfit$simple_fit))     # nr * nc: nothing is dropped
```

A **square** matrix is genuinely ambiguous — it could be one-mode, or
two-mode with equally large node sets — so it is assumed to be one-mode.
Say so explicitly when it is not:

```{r twomode-square}
sq <- list(y = yt[1:8, ], x1 = p1[1:8, ])
QAP(y ~ x1, data = sq, multi_mode = TRUE, reps = 50)$multi_mode
```

Groups for multi-mode data are given per dimension, as a *named* list:

```{r twomode-groups}
QAP(y ~ x1, data = list(y = yt, x1 = p1),
    groups = list(sender   = rep(1:3, each = 8),
                  receiver = rep(1:2, each = 4)),
    reps = 50)$abs["t", ]
```

Asking for something multi-mode data cannot support is an error rather than
a silent misapplication:

```{r twomode-errors, error = TRUE}
QAP(y ~ x1, data = list(y = yt, x1 = p1), mode = "undirected", reps = 10)
```

## Several networks at once

Pass a list of matrices (or arrays) per variable to pool across networks.
Each is permuted independently, and `random_intercept_nets` adds a
network-level random intercept.

```{r multinet}
mk <- function(k) matrix(rnorm(k^2), k, k)
xs <- list(mk(10), mk(12))
ys <- lapply(xs, function(x) 0.7 * x + matrix(rnorm(length(x)),
                                              nrow(x), ncol(x)))

QAP(y ~ x1, data = list(y = ys, x1 = xs), reps = 50)$coefficients
```

Group vectors can be given per network as an *unnamed* list, one element per
network — which is how they are told apart from the named per-dimension form
above.

## Building the input from a data frame

Dyadic data usually arrive as an edge list. `df_to_mat()` converts one to
the named list of matrices `QAP()` expects.

```{r df-to-mat}
edges <- expand.grid(sender = letters[1:6], receiver = letters[1:6],
                     stringsAsFactors = FALSE)
edges <- edges[edges$sender != edges$receiver, ]
edges$weight <- rnorm(nrow(edges))
edges$shared <- rnorm(nrow(edges))

mats <- df_to_mat(edges, sender = "sender", receiver = "receiver")
names(mats)

QAP(weight ~ shared, data = mats, reps = 50)$coefficients
```

It also handles a perceiver column (producing CSS arrays), `multi_mode` for
two-mode data, and `split_by` for several networks at once.

You can skip the conversion and hand the data frame straight to `QAP()`,
naming the dimension columns:

```{r df-direct, eval = FALSE}
QAP(y ~ x1, data = edges, sender = "from", receiver = "to", reps = 200)
```

`igraph` and `network` objects are accepted in the data list too, and are
converted to adjacency matrices; `edge_attr` names an edge attribute to
read as the value, rather than taking the binary adjacency.

## Longer runs

A p-value estimated from 200 permutations is granular to 1/200. Published
work normally wants 1000 or more; under `qapspp` that is 1000 *per
predictor*.

Two things help. `ncores` parallelises over permutations:

```{r parallel, eval = FALSE}
QAP(y ~ x1 + x2, data = d, reps = 5000, ncores = 4)
```

If you leave `ncores` unset, whatever `future` plan you have configured is
used unchanged, so you can manage parallelism yourself.

A long run can report progress. `QAP()` emits progress but does not decide
whether it is shown — wrap the call to switch it on:

```{r progress, eval = FALSE}
progressr::with_progress(
  QAP(y ~ x1 + x2, data = d, reps = 5000)
)
```

If memory is the constraint rather than time, `less_mem = TRUE` drops the
fitted model and the permutation draws, which is most of the object.
`coef()`, `nobs()` and `summary()` still work; `predict()`, `plot()` and
`confint()` do not.

And separate runs can be pooled afterwards, which is useful when a long job
has to be split:

```{r combine}
a <- QAP(y ~ x1 + x2, data = d, reps = 100, seed = 1)
b <- QAP(y ~ x1 + x2, data = d, reps = 100, seed = 2)

ab <- combine_qap_estimates(a, b)
ab$reps
```

Pooling is weighted by the permutations that actually contributed, so the
runs need not be the same length, and a run that lost permutations to
non-convergence is not over-weighted. The permutation draws are pooled too,
so `confint()` on the result uses all of them. For models fitted with
`comparison=`, each comparison is pooled separately.

Only repeated runs of the *same* model on the *same* data can be pooled,
and that is checked: the settings must match, and so must the coefficients,
which differ if and only if the data did.

## Notes on speed

Linear models with complete data take a fast path that precomputes
everything invariant across permutations and solves the normal equations
directly, rather than rebuilding the model frame and going through the
formula interface every time. It is roughly twelve times faster on network
matrices and fourteen on CSS arrays, and produces identical numbers — the
test suite checks the two paths against each other at a tolerance of 1e-12.

Under `qapspp` there is a further saving. Only one column of the design
changes per permutation, and only that column's coefficient and t-value are
ever compared with the baseline, so the fit is obtained by
Frisch–Waugh–Lovell from a decomposition of the columns that do *not*
change — formed once per predictor — instead of re-solving the whole design
each time. That is exact algebra rather than an approximation, and it puts
semi-partialling at roughly the same cost per permutation as outcome
permutation.

Anything the fast path cannot handle — a GLM, random effects, absorbed
fixed effects, comparisons, weights, an offset, or missing data —
automatically uses the general engine.

`use_gpu = TRUE` additionally offloads batched linear models to a GPU via
`torch`, if one is available:

```{r gpu, eval = FALSE}
gpu_available()
QAP(y ~ x1 + x2, data = d, reps = 10000, use_gpu = TRUE)
```

The GPU is worth it only on reasonably large networks: below a few thousand
dyads the transfer overhead exceeds the work, and on a CPU-only `torch`
install it is slower than the plain fast path throughout. Each run
re-checks its first batch against the CPU solver and stops if the two
disagree, so a broken driver or `torch` upgrade fails loudly rather than
moving p-values quietly.

## References

Dekker, D., Krackhardt, D., & Snijders, T. A. B. (2007). Sensitivity of
MRQAP tests to collinearity and autocorrelation conditions.
*Psychometrika*, 72(4), 563–581.

Krackhardt, D. (1988). Predicting with networks: Nonparametric multiple
regression analysis of dyadic data. *Social Networks*, 10(4), 359–381.
