---
title: "MRQAP for networks, cognitive social structures and multi-mode data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MRQAP for networks, cognitive social structures and multi-mode data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(MrQAP)
set.seed(20260825)
```

## Why permutation inference

A regression on dyadic data violates the independence assumption in an
obvious way. Every tie sent by actor *i* shares a sender; every tie received
by *j* shares a receiver. Observations are therefore correlated in a
structured way that ordinary standard errors do not know about, and the
resulting p-values are badly anti-conservative.

The quadratic assignment procedure sidesteps the problem instead of
modelling it. Permute the network — relabel the nodes, keeping the structure
intact — refit, and repeat. That gives an empirical null distribution for
each coefficient, built from data with the same dependence structure as the
original but with the hypothesised relationship destroyed. The p-value is
the proportion of permutations that produced a statistic at least as extreme
as the observed one.

`MrQAP` exposes one function, `QAP()`, for all of this.

## A one-mode network

We simulate a 20-node network whose tie strength depends on two dyadic
covariates.

```{r simulate}
n  <- 20
x1 <- matrix(rnorm(n^2), n, n)   # e.g. shared activity
x2 <- matrix(rnorm(n^2), n, n)   # e.g. geographic distance

y <- 1 + 0.8 * x1 - 0.4 * x2 + matrix(rnorm(n^2), n, n)

d <- list(y = y, x1 = x1, x2 = x2)
```

The left-hand side of the formula names the outcome in `data`; the
right-hand side names the predictors.

```{r fit-basic}
fit <- QAP(y ~ x1 + x2, data = d, reps = 200)
fit
```

The table gives, for each coefficient, the proportion of permutations whose
t-value was below the observed one, above it, and above it in absolute
value. The last is the two-sided p-value and is usually what you want.

The intercept is marked `*` because semi-partialling plus provides no test
for it — the procedure works by residualising each predictor on the others,
which has no analogue for a constant.

Coefficients come out where you would expect:

```{r coefs}
round(fit$coefficients, 3)
```

## Getting the results out

Printing is for reading; the extractor methods are for using.

```{r extractors}
coef(fit)

summary(fit)$coefficients          # a data frame, one row per term

as.data.frame(fit)                 # the same, for stacking across models
```

`summary()` prints the same report as the fit itself, with significance
codes added:

```{r summary}
summary(fit)
```

Note how a p-value of zero is shown. A permutation p-value is a proportion
out of `reps` draws, so its resolution is `1/reps` — with 200 permutations
the smallest value distinguishable from zero is 0.005. It is reported as
`<0.005`, not as the `<2e-16` that a conventional coefficient table would
print, because the permutation test cannot support that claim.

### Confidence intervals

`confint()` builds intervals from the *spread* of the permutation
distribution, so they reflect the same dyadic dependence the p-values do:

```{r confint}
confint(fit)
```

It is worth being explicit about what this is not. The permutation
distribution is generated under the null of no effect, so it is centred on
zero — its percentiles form an interval around zero and are not a
confidence interval at all. Only its standard deviation *s* is used, as
`estimate ± 1.96 * s`. That is a normal approximation: for a strongly
skewed coefficient, treat it as indicative.

Compare it with the model-based interval, which assumes the independence
MRQAP exists to avoid — usually narrower, and not consistent with the
reported p-value:

```{r confint-compare}
rbind(permutation = confint(fit)["x1", ],
      model_based = confint(fit$simple_fit)["x1", ])
```

`vcov()`, `fitted()`, `residuals()`, `logLik()`, `AIC()`, `BIC()` and
`nobs()` all describe the *baseline* fit, and carry the same caveat: they
know nothing about the permutation inference.

## Choosing the null hypothesis

Two procedures are available.

`nullhyp = "qapspp"` — the default — is Dekker, Krackhardt and Snijders'
semi-partialling plus. For each predictor in turn it regresses that
predictor on all the others, permutes the *residuals*, and refits. This is
the more trustworthy test when the predictors are correlated with each
other, which is the normal case. It costs `reps` permutations per predictor.

`nullhyp = "qapy"` permutes the outcome instead. It is cheaper — `reps`
permutations in total — but is the weaker test under collinearity.

```{r nullhyp}
qy <- QAP(y ~ x1 + x2, data = d, nullhyp = "qapy", reps = 200)

round(rbind(qapspp = fit$abs["t", ], qapy = qy$abs["t", ]), 3)
```

With a single predictor the two coincide, and `QAP()` silently uses `qapy`.

## Families beyond the linear model

Binary ties are the common case. Everything else about the call is the same.

```{r binomial}
eta <- -0.5 + 1.2 * x1
yb  <- matrix(rbinom(n^2, 1, 1 / (1 + exp(-eta))), n, n)

bfit <- QAP(y ~ x1 + x2, data = list(y = yb, x1 = x1, x2 = x2),
            family = "binomial", reps = 200)
round(bfit$coefficients, 3)
```

For binomial models `QAP()` also returns a confusion matrix built by drawing
from the predicted probabilities rather than by rounding them at 0.5, which
gives an honest picture of classification uncertainty:

```{r confusion}
bfit$confusion_matrix$accuracy
```

`family` also accepts `"poisson"`, `"negbin"`, `"zip"` and `"multinom"`.

## Robust standard errors

`use_robust_errors = TRUE` gives heteroskedasticity-consistent standard
errors: the closed-form HC3 for linear models, and the sandwich estimator
for GLMs.

```{r robust}
rfit <- QAP(y ~ x1 + x2, data = d, use_robust_errors = TRUE, reps = 200)
round(rbind(model_based = fit$t, robust = rfit$t), 2)
```

They are not available for mixed models, where `QAP()` warns and falls back
to model-based errors rather than reporting a number that is not what it
claims to be.

## Restricting permutations to groups

If the network has a known block structure — classrooms within a school,
departments within a firm — permuting freely across it destroys more than
the hypothesised effect. `groups` restricts each permutation to occur within
groups.

```{r groups}
grp <- rep(1:4, each = 5)
QAP(y ~ x1 + x2, data = d, groups = grp, reps = 200)$abs["t", ]
```

## Cognitive social structures

A CSS records each actor's *perception* of the whole network: a 3D array
indexed `[sender, receiver, perceiver]`. `QAP()` detects the extra dimension
and permutes accordingly — one node order applied across all three
dimensions, since sender, receiver and perceiver are the same people.

```{r css}
m  <- 8
a1 <- array(rnorm(m^3), dim = c(m, m, m))
ya <- 0.7 * a1 + array(rnorm(m^3), dim = c(m, m, m))

cfit <- QAP(y ~ x1, data = list(y = ya, x1 = a1), reps = 100)
round(cfit$coefficients, 3)
```

Perceiver-level heterogeneity can be absorbed with a random intercept:

```{r css-re, eval = requireNamespace("lme4", quietly = TRUE)}
QAP(y ~ x1, data = list(y = ya, x1 = a1),
    random_intercept_perceiver = TRUE, reps = 50)$coefficients
```

## Multi-mode data

In two-mode (affiliation) data the rows and columns are *different* kinds of
thing — people and organisations, say. There is no diagonal to exclude,
because an actor cannot have a tie to itself across node sets, and no
reciprocity, because there is no reverse dyad. The permutation must
therefore draw an independent node order for each dimension.

`QAP()` infers this when the dimensions differ in size:

```{r twomode}
nr <- 24; nc <- 8                      # 24 people, 8 organisations
p1 <- matrix(rnorm(nr * nc), nr, nc)
yt <- 0.6 * p1 + matrix(rnorm(nr * nc), nr, nc)

tfit <- QAP(y ~ x1, data = list(y = yt, x1 = p1), reps = 100)
tfit$multi_mode
nrow(model.frame(tfit$simple_fit))     # nr * nc: nothing is dropped
```

A **square** matrix is genuinely ambiguous — it could be one-mode, or
two-mode with equally large node sets — so it is assumed to be one-mode.
Say so explicitly when it is not:

```{r twomode-square}
sq <- list(y = yt[1:8, ], x1 = p1[1:8, ])
QAP(y ~ x1, data = sq, multi_mode = TRUE, reps = 50)$multi_mode
```

Groups for multi-mode data are given per dimension, as a *named* list:

```{r twomode-groups}
QAP(y ~ x1, data = list(y = yt, x1 = p1),
    groups = list(sender   = rep(1:3, each = 8),
                  receiver = rep(1:2, each = 4)),
    reps = 50)$abs["t", ]
```

Asking for something multi-mode data cannot support is an error rather than
a silent misapplication:

```{r twomode-errors, error = TRUE}
QAP(y ~ x1, data = list(y = yt, x1 = p1), mode = "undirected", reps = 10)
```

## Several networks at once

Pass a list of matrices (or arrays) per variable to pool across networks.
Each is permuted independently, and `random_intercept_nets` adds a
network-level random intercept.

```{r multinet}
mk <- function(k) matrix(rnorm(k^2), k, k)
xs <- list(mk(10), mk(12))
ys <- lapply(xs, function(x) 0.7 * x + matrix(rnorm(length(x)),
                                              nrow(x), ncol(x)))

QAP(y ~ x1, data = list(y = ys, x1 = xs), reps = 50)$coefficients
```

Group vectors can be given per network as an *unnamed* list, one element per
network — which is how they are told apart from the named per-dimension form
above.

## Building the input from a data frame

Dyadic data usually arrive as an edge list. `df_to_mat()` converts one to
the named list of matrices `QAP()` expects.

```{r df-to-mat}
edges <- expand.grid(sender = letters[1:6], receiver = letters[1:6],
                     stringsAsFactors = FALSE)
edges <- edges[edges$sender != edges$receiver, ]
edges$weight <- rnorm(nrow(edges))
edges$shared <- rnorm(nrow(edges))

mats <- df_to_mat(edges, sender = "sender", receiver = "receiver")
names(mats)

QAP(weight ~ shared, data = mats, reps = 50)$coefficients
```

It also handles a perceiver column (producing CSS arrays), `multi_mode` for
two-mode data, and `split_by` for several networks at once.

## Longer runs

A p-value estimated from 200 permutations is granular to 1/200. Published
work normally wants 1000 or more; under `qapspp` that is 1000 *per
predictor*.

Two things help. `ncores` parallelises over permutations:

```{r parallel, eval = FALSE}
QAP(y ~ x1 + x2, data = d, reps = 5000, ncores = 4)
```

If you leave `ncores` unset, whatever `future` plan you have configured is
used unchanged, so you can manage parallelism yourself.

And separate runs can be pooled afterwards, which is useful when a long job
has to be split:

```{r combine}
a <- QAP(y ~ x1 + x2, data = d, reps = 100, seed = 1)
b <- QAP(y ~ x1 + x2, data = d, reps = 100, seed = 2)

ab <- combine_qap_estimates(a, b)
ab$reps
```

Pooling is weighted by `reps`, so the runs need not be the same length. For
models fitted with `comparison=`, each comparison is pooled separately.

## Notes on speed

Linear models with complete data take a fast path that precomputes
everything invariant across permutations and solves the normal equations
directly, rather than rebuilding the model frame and going through the
formula interface every time. It is roughly twelve times faster on network
matrices and fourteen on CSS arrays, and produces identical numbers — the
test suite checks the two paths against each other at a tolerance of 1e-12.

Anything the fast path cannot handle — a GLM, random effects, absorbed
fixed effects, comparisons, or missing data — automatically uses the general
engine.

`use_gpu = TRUE` additionally offloads batched linear models to a GPU via
`torch`, if one is available:

```{r gpu, eval = FALSE}
gpu_available()
QAP(y ~ x1 + x2, data = d, reps = 10000, use_gpu = TRUE)
```

## References

Dekker, D., Krackhardt, D., & Snijders, T. A. B. (2007). Sensitivity of
MRQAP tests to collinearity and autocorrelation conditions.
*Psychometrika*, 72(4), 563–581.

Krackhardt, D. (1988). Predicting with networks: Nonparametric multiple
regression analysis of dyadic data. *Social Networks*, 10(4), 359–381.
